---
title: "Ch7_Interactions"
author: "Michael Rose"
date: "June 23, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rethinking)
```

To model deeper conditionality - where the importance of one predictor depends upon another predictor - we need **interaction**. 
Interaction is a kind of conditioning, a way to allow parameters to be conditional on further aspects of the data. 

# 7.1 | Building an Interaction 

```{r}
# load data
data(rugged)
d <- rugged

# make log version of outcome 
d$log_gdp <- log(d$rgdppc_2000)

# extract countries with GDP data 
dd <- d[complete.cases(d$rgdppc_2000),]

# split countries into Africa and not-Africa
d.A1 <- dd[dd$cont_africa == 1,] # Africa
d.A0 <- dd[dd$cont_africa == 0,] # Not Africa

```

Our outcome variable here is the log of real gross domestic product per capita from the year 2000, rgdpoc_2000. The log is the magnitude of GDP. 

```{r}
# African Nations 
m7.1 <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma), 
    mu <- a + bR * rugged, 
    a ~ dnorm(8, 100), 
    bR ~ dnorm(0, 1), 
    sigma ~ dunif(0, 10)
  ), data = d.A1
)

# Non African Nations 
m7.2 <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma), 
    mu <- a + bR * rugged, 
    a ~ dnorm(8, 100), 
    bR ~ dnorm(0, 1), 
    sigma ~ dunif(0, 10)
  ), data = d.A0
)

# plot African Nations 
plot(log_gdp ~ rugged, data = d.A1, col = rangi2)
abline(m7.1)

# plot non African Nations 
plot(log_gdp ~ rugged, data = d.A0, col = rangi2)
abline(m7.2)
```

## 7.1.1 | Adding a dummy variable doesn't work 

We are fitting this to show that adding a dummy variable for cont_africa doesn't reverse the slope of the linear regression. 

```{r}
# fit model without dummy variable 
m7.3 <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR * rugged, 
    a ~ dnorm(8, 100), 
    bR ~ dnorm(0, 1), 
    sigma ~ dunif(0, 10)
  ), data = dd
)

# fit model with dummy variable 
m7.4 <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma), 
    mu <- a + bR * rugged + bA * cont_africa, 
    a ~ dnorm(8, 100), 
    bR ~ dnorm(0, 1), 
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ), data = dd
)

# compare models using WAIC 
compare(m7.3, m7.4)
```

We can ignore m7.3, since m7.4 gets 100% of the model weight. Now we will plot the posterior predictions from m7.4 so we can see how, despite its plausible superiority to m7.3, it still doesn't manage different slopes inside and outside of africa. 

```{r}
# sample from the posterior and compute the predicted means and intervals for both african and non african nations 

# set sequence 
rugged.seq <- seq(from = -1, to = 8, by = 0.25)

# compute mu over samples, fixing cont_africa = 0
mu.NotAfrica <- link(m7.4, data = data.frame(cont_africa = 0, rugged = rugged.seq))

# compute mu over samples, fixing cont_africa = 1
mu.Africa <- link(m7.4, data = data.frame(cont_africa = 1, rugged = rugged.seq))

# summarize to means and intervals 
mu.NotAfrica.mean <- apply(mu.NotAfrica, 2, mean)
mu.Africa.mean <- apply(mu.Africa, 2, mean)
mu.NotAfrica.PI <- apply(mu.NotAfrica, 2, PI, prob = 0.97)
mu.Africa.PI <- apply(mu.Africa, 2, PI, prob = 0.97)

# plot Africa
plot(log_gdp ~ rugged , data=d.A1)
lines( rugged.seq , mu.Africa.mean )
shade( mu.Africa.PI, rugged.seq )

# plot not Africa and Africa together 
plot(log_gdp ~ rugged, data = d.A0)
points(log_gdp ~ rugged, data = d.A1, col = rangi2)
lines(rugged.seq, mu.NotAfrica.mean)
shade(mu.NotAfrica.PI, rugged.seq)
lines( rugged.seq , mu.Africa.mean, col = rangi2)
shade( mu.Africa.PI, rugged.seq)
```

### 7.1.2 | Adding a Linear Interaction does work 

To recover the change in slope we say at the start of the section, we need to add a proper interaction effect. 

We will do this by adding a relationship between a slope and a predictor with a linear model, e.g. : 

$Y_i \sim \mathrm{Normal}(\mu_i, \sigma)$        # Likelihood Function 
$\mu_i = \alpha + \gamma_i R_i + \beta_A A_i$    # Linear Model of $\mu$
$\gamma_i = \beta_R + \beta_{AR} A_i$            # Linear model of slope 

Where R_i is ruggedness and A_i is african nation 

The equation for $\gamma_i$ defines the interaction between ruggedness and African nations. It is a linear interaction effect, since gamma is a linear model. By defining the relationship between GDP and ruggedness this way, we are explicitly modeling the hypothesis that the slope between GDP and ruggedness depends - is conditional - upon whether or not a nation is in Africa. 
Therefore: 
  if $\beta_{AR}$ = 0, we get our previous model back and African Nations has no effect
  if $\beta_{AR}$ > 0, then African nations has a positive effect on the slope between gdp and ruggedness 
  if $\beta_{AR}$ < 0, then African nations has a negative effect on the slope between gdp and ruggedness 

```{r}
# fit model with interaction 
m7.5 <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma), 
    mu <- a + gamma * rugged + bA * cont_africa,
    gamma <- bR + bAR * cont_africa, 
    a ~ dnorm(8, 100), 
    bA ~ dnorm(0, 1),
    bR ~ dnorm(0, 1), 
    bAR ~ dnorm(0, 1), 
    sigma ~ dunif(0, 10)
  ), data = dd
)

# use WAIC to compare this model with the other two 
compare(m7.3, m7.4, m7.5)

```

#### Overthinking | Conventional form of interaction 

The method above is identical to the common form of defining interactions, 

$Y_i \sim \mathrm{Normal}(\mu_i, \sigma)$
$\mu+i = \alpha + \beta_R R_i + \beta_{AR} A_iR_i + \beta_A A_i$

The equation for $\gamma$ has just been substituted into the second line and expanded. 

```{r}
# expanded form
m7.5b <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma), 
    mu <- a + bR * rugged + bAR * rugged * cont_africa + bA * cont_africa, 
    a ~ dnorm(8, 100), 
    bA ~ dnorm(0, 1), 
    bR ~ dnorm(0, 1), 
    bAR ~ dnorm(0, 1), 
    sigma ~ dunif(0, 10)
  ), data = dd
)

```


## 7.1.3 | Plotting the Interaction 

```{r}
# calculate the posterior mean line and interval 
rugged.seq <- seq(from = -1, to = 8, by = 0.25)

mu.Africa <- link(m7.5, data = data.frame(cont_africa = 1, rugged = rugged.seq)) 
mu.Africa.mean <- apply(mu.Africa, 2, mean)
mu.Africa.PI <- apply(mu.Africa, 2, PI, prob = 0.97) 

mu.NotAfrica <- link(m7.5, data = data.frame(cont_africa = 0, rugged = rugged.seq))
mu.NotAfrica.mean <- apply(mu.NotAfrica, 2, mean)
mu.NotAfrica.PI <- apply(mu.NotAfrica, 2, PI, prob = 0.97)

# plot 

# African Nations with Regression 
d.A1 <- dd[dd$cont_africa == 1,]
plot(log(rgdppc_2000) ~ rugged, data = d.A1, col = rangi2, ylab = "log GDP year 2000", xlab = "Terrain Ruggedness Index")
mtext("African Nations", 3) 
lines(rugged.seq, mu.Africa.mean, col = rangi2) 
shade(mu.Africa.PI, rugged.seq, col = col.alpha(rangi2, 0.3))

# Non African Nations with Regression 
d.A0 <- dd[dd$cont_africa == 0,]
plot(log(rgdppc_2000) ~ rugged, data = d.A0, col = rangi2, ylab = "log GDP year 2000", xlab = "Terrain Ruggedness Index")
mtext("Non African Nations", 3)
lines(rugged.seq, mu.NotAfrica.mean)
shade(mu.NotAfrica.PI, rugged.seq)
```

## 7.1.4 | Interpreting an Interaction Estimate 

Generally, we do not want to rely on a table of numbers to interpret a model with an interaction. 

### 7.1.4.1 | Parameters Change Meaning 

When we add an interaction term to a model, the slope that the interaction term is predicting is now reliant on every parameter involved in the interaction. As a result, in order to interpret the slope of the interaction term, we need to interpret each parameter involved with the interaction term. 

As a result, we can no longer read the influence of either predictor from a table of elements. 

```{r}
precis(m7.5)
```

As we can see, the table doesn't even show $\gamma$, we would need to calculate it ourselves. 

### 7.1.4.2 | Incorporating Uncertainty 

Since $\gamma$ depends upon parameters, and those parameters have posterior distributions, $\gamma$ must also have a posterior distribution. 

```{r}
# compute posterior distribution of gamma 
post <- extract.samples(m7.5)
gamma.Africa <- post$bR + post$bAR * 1 
gamma.notAfrica <- post$bR + post$bAR * 0

# calculate means 
mean(gamma.Africa)
mean(gamma.notAfrica)

# plot 
dens(gamma.Africa, xlim = c(-0.5, 0.6), ylim = c(0, 5.5), xlab = "gamma", col = rangi2)
dens(gamma.notAfrica, add = TRUE)

```

We can see the posterior distributions for the effect that ruggedness has on the gdp of african and non african countries above. We see that generally, ruggedness has a positive effect on african nations and a negative effect on non african nations. 

```{r}
# calculate the difference between the slopes
diff <- gamma.Africa - gamma.notAfrica

# check what proportion of these differences is below zero 
sum(diff < 0) / length(diff)

```

# 7.2 | Symmetry of the Linear Interaction 

Due to symmetry, our golem doesn't see a difference between the following 2 questions: 
  1. How much does the influence of ruggedness (on GDP) depend upon whether the nation is in Africa? 
  2. How much does the influence of being in Africa (on GDP) depend upon ruggedness? 

## 7.2.2 | Africa depends upon ruggedness 

```{r}
# plot The influence of being in Africa depends upon terrain ruggedness 

# get minimum and maximum rugged values 
q.rugged <- range(dd$rugged)

# compute lines and confidence intervals 
mu.ruggedlo <- link(m7.5, data = data.frame(rugged = q.rugged[1], cont_africa = 0:1))
mu.ruggedlo.mean <- apply(mu.ruggedlo, 2, mean)
mu.ruggedlo.PI <- apply(mu.ruggedlo, 2, PI)

mu.ruggedhi <- link(m7.5, data = data.frame(rugged = q.rugged[2], cont_africa = 0:1))
mu.ruggedhi.mean <- apply(mu.ruggedhi, 2, mean)
mu.ruggedhi.PI <- apply(mu.ruggedhi, 2, PI)

# plot, splitting points at median 

# get median 
med.r <- median(dd$rugged)

# 
ox <- ifelse(dd$rugged > med.r, 0.05, -0.05)

# plot 
plot(dd$cont_africa + ox, log(dd$rgdppc_2000), col = ifelse(dd$rugged > med.r, rangi2, "black"), 
     xlim = c(-0.25, 1.25), xaxt = "n", ylab = "log GDP year 2000", xlab = "Continent")
axis(1, at = c(0, 1), labels = c("other", "Africa"))
lines(0:1, mu.ruggedlo.mean, lty = 2)
shade(mu.ruggedlo.PI, 0:1)
lines(0:1, mu.ruggedhi.mean, col = rangi2)
shade(mu.ruggedhi.PI, 0:1, col = col.alpha(rangi2, 0.25))

```

In the graph above, black points are ones below the median in ruggedness, and blue are the countries above the median in ruggedness.

# 7.3 | Continuous Interactions 

```{r}
data(tulips)
d <- tulips
str(d)
```

```{r}
# fit models
m7.6 <- map(
  alist(
    blooms ~ dnorm(mu, sigma), 
    mu <- a + bW * water + bS * shade, 
    a ~ dnorm(0, 100), 
    bW ~ dnorm(0, 100), 
    bS ~ dnorm(0, 100), 
    sigma ~ dunif(0, 100)
  ), data = d, method = "Nelder-Mead", control = list(maxit = 1e4)
)

m7.7 <- map(
  alist(
    blooms ~ dnorm(mu, sigma), 
    mu <- a + bW * water + bS * shade + bWS * water * shade, 
    a ~ dnorm(0, 100), 
    bW ~ dnorm(0, 100), 
    bS ~ dnorm(0, 100), 
    bWS ~ dnorm(0, 100), 
    sigma ~ dunif(0, 100)
  ), data = d, method = "Nelder-Mead", control = list(maxit = 1e4)
)

# look at estimates 
coeftab(m7.6, m7.7)

# compare 
compare(m7.6, m7.7)
```

If we get an error that our optimization search space is too large, we can fix it with 3 basic solutions: 
  1. Use another method of optimization. R's optim knows several. map uses BFGS, but we can also try Nelder-Mean and SANN 
  2. We can tell optim to search longer so it doesn't reach maximum iterations 
  3. We can rescale the data, so that the search has less terrain to search 
  
## 7.3.3 | Center and Re-estimate 

```{r}
# subtract mean of the original from each value 
d$shade.c <- d$shade - mean(d$shade)
d$water.c <- d$water - mean(d$water)
```

centering will do two things for us: 
  1. It will fix the problem of convergence in optimization
  2. It will make our estimates easier to interpret 
  
```{r}
m7.8 <- map(
  alist(
    blooms ~ dnorm(mu, sigma), 
    mu <- a + bW * water.c + bS * shade.c, 
    a ~ dnorm(0, 100), 
    bW ~ dnorm(0, 100), 
    bS ~ dnorm(0, 100), 
    sigma ~ dunif(0, 100)
  ), data = d, start = list(a = mean(d$blooms), bW = 0, bS = 0, sigma = sd(d$blooms))
)

m7.9 <- map(
  alist(
    blooms ~ dnorm(mu, sigma), 
    mu <- a + bW * water.c + bS * shade.c + bWS * water.c * shade.c, 
    a ~ dnorm(130, 100), 
    bW ~ dnorm(0, 100), 
    bS ~ dnorm(0, 100), 
    bWS ~ dnorm(0, 100), 
    sigma ~ dunif(0, 100)
  ), data = d, start = list(a = mean(d$blooms), bW = 0, bS = 0, bWS = 0, sigma = sd(d$blooms))
)

coeftab(m7.8, m7.9)

```

```{r}
# compute the effect of increasing water by 1 unit 

# regular interaction 
k <- coef(m7.7)
k[1] + k[2] * 2 + k[3] * 2 + k[4] * 2 * 2

# centered interaction 
k <- coef(m7.9)
k[1] + k[2] * 0 + k[3] * 0 + k[4] * 0 * 0

# centered estimates 
precis(m7.9)
```

 - The estimate a is the expected value of blooms when both water and shade are at their average values 
 - The estimate bW is the expected change in blooms when water increases by one unit *and* shade is at its average value 
 - The estimate bS is the expected change in blooms when shade increases by one unit *and* water is at its average value 
 - The estimate bWS is the interaction effect. This tells us two things: 
    - The expected change in the influence of water on blooms when increasing shade by one unit 
    - The expected change in the influence of shade on blooms when increasing water by one unit 
    
## 7.3.4 | Plotting Implied Predictions 

```{r}
# Non Interaction Model 

# make a plot window with three panels in a single row 
par(mfrow = c(1, 3)) # 1 row, 3 columns 

# loop over values of water.c and plot predictions 
shade.seq <- -1:1 
max(d$blooms)
d$bl

for (w in -1:1){
  dt <- d[d$water.w == w,]
  plot( blooms ~ shade.c , data=dt , col=rangi2 , main=paste("water.c =",w) , xlim = c(-1,1) , ylim=c(0,362), xlab="shade (centered)" )
  mu <- link(m7.8, data = data.frame(water.c = w, shade.c = shade.seq))
  mu.mean <- apply(mu, 2, mean)
  mu.PI <- apply(mu, 2, PI, prob = 0.97)
  lines(shade.seq, mu.mean)
  lines(shade.seq, mu.PI[1,], lty = 2)
  lines(shade.seq, mu.PI[2,], lty = 2)
}
```


```{r}

# Interaction Model 

# make a plot window with three panels in a single row 
par(mfrow = c(1, 3)) # 1 row, 3 columns 

# loop over values of water.c and plot predictions 
shade.seq <- -1:1 
max(d$blooms)
d$bl

for (w in -1:1){
  dt <- d[d$water.w == w,]
  plot( blooms ~ shade.c , data=dt , col=rangi2 , main=paste("water.c =",w) , xlim = c(-1,1) , ylim=c(0,362), xlab="shade (centered)" )
  mu <- link(m7.9, data = data.frame(water.c = w, shade.c = shade.seq))
  mu.mean <- apply(mu, 2, mean)
  mu.PI <- apply(mu, 2, PI, prob = 0.97)
  lines(shade.seq, mu.mean)
  lines(shade.seq, mu.PI[1,], lty = 2)
  lines(shade.seq, mu.PI[2,], lty = 2)
}

```

# 7.4 | Interactions in Design Formulas 

Suppose we want the following model: 

$y_i \sim \mathrm{Normal}(\mu_i, \sigma)$ 
$\mu_i = \alpha + \beta_x x_i + \beta_z z_i + \beta_{xz}x_iz_i$ 

```{DNR}
# as a design formula 
m7.x <- lm(y ~ x + z + x*z, data = d)

# we can get the full model using lm with just this 
m7.x <- lm(y ~ x*z, data = d)

# if we don't want the main effects, we must subtract them 
m7.x <- lm(y ~ x + x*z - z, data = d)

# if we wanted a huge model with x, z, w terms and all their interactions 
m7.x <- lm(y ~ x*z*w, data = d)

```


